Peak area control doc

### 2023-05-25 targeted/targeted_control.Rmd
Updating targeted to modern syntax

clean_stans <- read.csv("../metadata/made_data/clean_stans.csv")
name_conversion <- clean_stans %>%
  select(new=compound_name, old=compound_name_old) %>%
  distinct()
file_metadata <- read.csv("../metadata/made_data/file_metadata.csv")
```

@@

```{r MSDIAL autoexport things not yet}
# Not implemented yet, check on thread below for updates
# http://www.metabolomics-forum.com/index.php?topic=1820.0
# Instead, open an existing .mtd2 file, click the "Export" tab, select "Alignment result", specify the output directory as "Area_HILICPos_EddyTransect.csv", check a couple of the boxes (Raw Data Area is the only critical one), change the output format to msp(?), and hit the "Export" button at the bottom. You then need to open this file in Excel and save as CSV.

# Haven't been able to find the raw data for negative mode so I'm currently just copying this off the Google Drive (G:\Shared drives\Ingalls Lab\Collaborative Projects\MESO-SCOPE\EddyTransect\HILIC_Neg_MSDial)
```

@@

```{r transectSkylineImport}
manual_pos <- read_csv("raw_data/Mesotransect/Area_HILICPos_EddyTransect.csv", skip = 4) %>%
  filter(`Metabolite name`!="Unknown") %>%
  select(compound_name=`Metabolite name`, starts_with("180821")) %>%
  pivot_longer(cols=starts_with("180821"), names_to = "filename", values_to="area") %>%
  mutate(polarity="pos")
manual_neg <- read_csv("raw_data/Mesotransect/Area_HILICNeg_EddyTransect.csv", skip = 4) %>%
  filter(`Metabolite name`!="Unknown") %>%
  select(compound_name=`Metabolite name`, starts_with("180821")) %>%
  pivot_longer(cols=starts_with("180821"), names_to = "filename", values_to="area") %>%
  mutate(polarity="neg")
manual_ints <- rbind(manual_pos, manual_neg) %>%
  mutate(compound_name=str_replace_all(compound_name, "Ingalls_|w/o MS2:|Unsettled: ", "")) %>%
  mutate(filename = ifelse(str_detect(filename, "DDA"),
                           str_replace(filename, "Poo", "DDA"),
                           filename)) %>%
  mutate(filename = ifelse(str_detect(filename, "QC"),
                           str_replace(filename, "Poo", "QC"),
                           filename))

# Once again, filter only those that Raafay integrated
clean_ints_transect <- manual_ints %>%
  filter(!str_detect(filename, "DDA")) %>%
  filter(compound_name!="Methyl 3-methylthiopropionate") %>%
  mutate(compound_name=str_replace(compound_name, "Monophosphate, 15N5", "monophosphate, 15N5")) %>%
  left_join(name_conversion, by=c("compound_name"="old")) %>% 
  mutate(compound_name=new) %>% 
  select(-new) %>% 
  left_join(manual_quality, by=c("compound_name", "polarity")) %>%
  filter(integrated=="Y" &in_samps%in%c("Y", "Maybe?", "Kinda?", "Barely", "Kinda")) %>%
  select(-integrated, -in_samps) %>%
  mutate(area=as.numeric(area)) %>%
  anti_join(distinct(., compound_name, polarity) %>% filter(duplicated(compound_name)))

```

@@

mesotransect_data <- rfs_transect %>% 
  left_join(BMISed_transect, by="compound_name") %>%
  left_join(clean_stans) %>%
  mutate(uM = BMISed_area/rf_h2o * 0.0004/10 * 1/rf_ratio * 2) %>% # 400uL vial, 10L sample, diluted 2x
  mutate(nM=uM*1000) %>%
  select(compound_name, nM, filename, polarity)
```

@@

## Write out quant data
quant_data <- bind_rows(falkor_data, mesotransect_data, mesocenter_data) %>%
  filter(compound_name!="Choline")
write.csv(quant_data, file = "made_data/longdata.csv", row.names = FALSE)
```

```{r write MESOSCOPE to Excel}
library(openxlsx)
wb <- createWorkbook(creator = "William Kumler, wkumler@uw.edu")
addWorksheet(wb, sheetName = "Sample data")
bind_rows(mesotransect_data, mesocenter_data) %>%
  pivot_wider(names_from = filename, values_from = nM) %>%
  writeData(wb = wb, sheet = "Sample data")
freezePane(wb, sheet = "Sample data", firstRow = TRUE, firstCol = TRUE)
addWorksheet(wb, "File metadata")
file_metadata %>%
  filter(cruise%in%c("MT", "MC")) %>%
  mutate(filename=str_remove(filename, ".mzML")) %>%
  distinct(filename, samp_type, station, cast, depth, abs_depth, rosette,date_run, timestamp) %>%
  rename(sample_name=filename, file_timestamp=timestamp) %>%
  writeData(wb = wb, sheet = "File metadata")
addWorksheet(wb = wb, sheetName = "Compound metadata")
clean_stans %>%
  select(compound_name, compound_type, formula, rt, mz, kegg_id, polarity, peak_quality) %>%
  writeData(wb = wb, sheet = "Compound metadata")

addWorksheet(wb=wb, sheetName = "README")
data.frame(`README for Ingalls MESOSCOPE data`=c(
  "This file was produced by William Kumler in the Ingalls lab at the University of Washington",
  "This file contains concentrations of compounds in seawater measured via HPLC-MS",
  "in a targeted (selected subset of known compounds) method.",
  "",
  "The first sheet (Sample data) contains the concentrations of various molecules in each sample",
  "taken during the MESOSCOPE cruise in 2017. These values are in nanomoles per",
  "liter of seawater (nM) from the particulate phase (particles collected onto",
  "0.2um Durapore filters).",
  "Missing values are shown as empty cells and infinite values (usually in the standards) are shown as #NUM!",
  "",
  "The second sheet (File metadata) contains information about the samples themselves, including",
  "the depth, time, and station from which they were collected. ",
  "sample_name = Name of the sample as shown in the Sample data Excel sheet",
  "samp_type = Type of sample, whether biological sample (Smp), blank (Blk), pooled (Poo), or standard (Std)",
  "station = Station from which the sample was collected, 1-15 plus L1 and L2",
  "cast = Which cast the sample was taken from",
  "depth = Depth descriptor of where in the water column the sample was collected",
  "abs_depth = Absolute depth. Accounts for differences from DCM and is determined from the CTD itself",
  "rosette = Which Niskin was this sample taken from? Largely incomplete",
  "date_run = The date on which the HPLC-MS run was started",
  "file_timestamp = The time at which the sample was actually run",
  "",
  "The third sheet (Compound metadata) contains information about the compounds measured",
  "during this analysis. Not all compounds were detected and passed quality control",
  "so there are more compounds here than found in the Sample data sheet.",
  "compound_name = Name of the compound, as found in the Sample data sheet",
  "compound_type = Broad compound class as determined rather arbitrarily by myself",
  "formula = Molecular formula of the compound. Isotopes are labeled in parentheses",
  "rt = Estimated retention time of the compound, in minutes",
  "mz = Mass-to-charge ratio of the molecule at which it's detected in the mass spec",
  "kegg_id = KEGG (Kyoto Encyclopedia of Genes and Genomes) ID: https://www.genome.jp/kegg/",
  "polarity = The polarity at which the molecule was detected in the mass spec, either 'pos' for positive mode or 'neg' for negative mode",
  "peak_quality = The typical quality of the chromatographic peak. Yes is good, No is bad, and see the Ingalls Standards repo for more details on Problematic compounds",
  "",
  paste("File created on", Sys.Date())
)) %>%
  writeData(wb = wb, sheet = "README")
saveWorkbook(wb, file = "MESOSCOPE_Targeted_nM.xlsx", overwrite = TRUE)
```

```{r write Falkor to Excel}
library(openxlsx)
wb <- createWorkbook(creator = "William Kumler, wkumler@uw.edu")
addWorksheet(wb, sheetName = "Sample data")
falkor_data %>%
  pivot_wider(names_from = filename, values_from = nM) %>%
  writeData(wb = wb, sheet = "Sample data")
freezePane(wb, sheet = "Sample data", firstRow = TRUE, firstCol = TRUE)
addWorksheet(wb, "File metadata")
file_metadata %>%
  filter(cruise=="FK") %>%
  mutate(filename=str_remove(filename, ".mzML")) %>%
  distinct(filename, samp_type, station, cast, depth, abs_depth, rosette, date_run, timestamp) %>%
  rename(sample_name=filename, file_timestamp=timestamp) %>%
  writeData(wb = wb, sheet = "File metadata")
addWorksheet(wb = wb, sheetName = "Compound metadata")
clean_stans %>%
  select(compound_name, compound_type, formula, rt, mz, kegg_id, polarity, peak_quality) %>%
  writeData(wb = wb, sheet = "Compound metadata")

addWorksheet(wb=wb, sheetName = "README")
data.frame(`README for Ingalls Falkor data`=c(
  "This file was produced by William Kumler in the Ingalls lab at the University of Washington",
  "This file contains concentrations of compounds in seawater measured via HPLC-MS",
  "in a targeted (selected subset of known compounds, manually integrated) method.",
  "",
  "The first sheet (Sample data) contains the concentrations of various molecules in each sample",
  "taken during the Falkor cruise in 2018. These values are in nanomoles per",
  "liter of seawater (nM) from the particulate phase (particles collected onto",
  "0.2um Durapore filters).",
  "Missing values are shown as empty cells and infinite values (usually in the standards) are shown as #NUM!",
  "",
  "The second sheet (File metadata) contains information about the samples themselves, including",
  "the depth, time, and station from which they were collected. ",
  "sample_name = Name of the sample as shown in the Sample data Excel sheet",
  "samp_type = Type of sample, whether biological sample (Smp), blank (Blk), pooled (Poo), or standard (Std)",
  "station = Station from which the sample was collected",
  "cast = Which cast the sample was taken from",
  "depth = Depth descriptor of where in the water column the sample was collected",
  "abs_depth = Absolute depth. Accounts for differences from DCM and is determined from the CTD itself",
  "rosette = Which Niskin was this sample taken from? Largely incomplete",
  "date_run = The date on which the HPLC-MS run was started",
  "file_timestamp = The time at which the sample was actually run",
  "",
  "The third sheet (Compound metadata) contains information about the compounds measured",
  "during this analysis. Not all compounds were detected and passed quality control",
  "so there are more compounds here than found in the Sample data sheet.",
  "compound_name = Name of the compound, as found in the Sample data sheet",
  "compound_type = Broad compound class as determined rather arbitrarily by myself",
  "formula = Molecular formula of the compound. Isotopes are labeled in parentheses",
  "rt = Estimated retention time of the compound, in minutes",
  "mz = Mass-to-charge ratio of the molecule at which it's detected in the mass spec",
  "kegg_id = KEGG (Kyoto Encyclopedia of Genes and Genomes) ID: https://www.genome.jp/kegg/",
  "polarity = The polarity at which the molecule was detected in the mass spec, either 'pos' for positive mode or 'neg' for negative mode",
  "peak_quality = The typical quality of the chromatographic peak. Yes is good, No is bad, and see the Ingalls Standards repo for more details on Problematic compounds",
  "",
  paste("File created on", Sys.Date())
)) %>%
  writeData(wb = wb, sheet = "README")
saveWorkbook(wb, file = "Falkor_Targeted_nM.xlsx", overwrite = TRUE)
```

>>>>

clean_stans <- read.csv("../metadata/clean_stans.csv")
name_conversion <- clean_stans %>%
  select(new=compound_name, old=name_old) %>%
  distinct()
file_metadata <- read.csv("../metadata/filled_file_metadata.csv")
```

@@

```{r Mesotransect MSDIAL autoexport things not yet}
# Not implemented yet, check on thread below for updates
# http://www.metabolomics-forum.com/index.php?topic=1820.0
# Instead, open an existing .mtd2 file, click the "Export" tab, select "Alignment result", specify the output directory as the current working directory, check the Raw Data matrix (Area) box , keep the output format as msp, and hit the "Export" button at the bottom.
```

@@

```{r transectMsdialImport}

manual_neg <- read_delim("raw_data/Mesotransect/MSDial_neg_data/Area_0_20235251247.txt", skip = 4, delim = "\t") %>%
  filter(`Metabolite name`!="Unknown") %>%
  select(compound_name=`Metabolite name`, starts_with("180821")) %>%
  pivot_longer(cols=starts_with("180821"), names_to = "filename", values_to="area") %>%
  mutate(polarity="neg")
manual_pos <- read_delim("raw_data/Mesotransect/MSDial_pos_data/Area_2_2023525133.txt", skip = 4, delim = "\t") %>%
  filter(`Metabolite name`!="Unknown") %>%
  select(compound_name=`Metabolite name`, starts_with("180821")) %>%
  pivot_longer(cols=starts_with("180821"), names_to = "filename", values_to="area") %>%
  mutate(polarity="pos")
manual_ints <- rbind(manual_pos, manual_neg) %>%
  mutate(compound_name=str_replace_all(compound_name, "Ingalls_|w/o MS2:|Unsettled: ", "")) %>%
  mutate(filename = ifelse(str_detect(filename, "DDA"),
                           str_replace(filename, "Poo", "DDA"),
                           filename)) %>%
  mutate(filename = ifelse(str_detect(filename, "QC"),
                           str_replace(filename, "Poo", "QC"),
                           filename))

# Once again, filter only those that Raafay integrated
clean_ints_transect <- manual_ints %>%
  filter(!str_detect(filename, "DDA")) %>%
  filter(compound_name!="Methyl 3-methylthiopropionate") %>%
  mutate(compound_name=str_replace(compound_name, "Monophosphate, 15N5", "monophosphate, 15N5")) %>%
  left_join(name_conversion, by=c("compound_name"="old")) %>% 
  mutate(compound_name=ifelse(is.na(new), compound_name, new)) %>% 
  select(-new) %>% 
  left_join(manual_quality, by=c("compound_name", "polarity")) %>%
  filter(integrated=="Y" &in_samps%in%c("Y", "Maybe?", "Kinda?", "Barely", "Kinda")) %>%
  select(-integrated, -in_samps) %>%
  mutate(area=as.numeric(area)) %>%
  anti_join(distinct(., compound_name, polarity) %>% filter(duplicated(compound_name)))
```

@@

mesotransect_data <- rfs_transect %>% 
  left_join(BMISed_transect, by="compound_name", multiple="all") %>%
  left_join(clean_stans) %>%
  mutate(uM = BMISed_area/rf_h2o * 0.0004/10 * 1/rf_ratio * 2) %>% # 400uL vial, 10L sample, diluted 2x
  mutate(nM=uM*1000) %>%
  select(compound_name, nM, filename, polarity)
```

@@

## Write out quant data
quant_data <- bind_rows(falkor_data, mesotransect_data, mesocenter_data) %>%
  filter(compound_name!="Choline")
write.csv(quant_data, file = "made_data/longdata.csv", row.names = FALSE)
write_csv(quant_data, file = "made_data/all_peaks.csv")
```

```{r write MESOSCOPE to Excel}
library(openxlsx)
wb <- createWorkbook(creator = "William Kumler, wkumler@uw.edu")
addWorksheet(wb, sheetName = "Sample data")
bind_rows(mesotransect_data, mesocenter_data) %>%
  pivot_wider(names_from = filename, values_from = nM) %>%
  writeData(wb = wb, sheet = "Sample data")
freezePane(wb, sheet = "Sample data", firstRow = TRUE, firstCol = TRUE)
addWorksheet(wb, "File metadata")
file_metadata %>%
  filter(cruise%in%c("MT", "MC")) %>%
  mutate(filename=str_remove(filename, ".mzML")) %>%
  distinct(filename, samp_type, station, cast, depth, abs_depth, rosette,date_run) %>%
  rename(sample_name=filename) %>%
  writeData(wb = wb, sheet = "File metadata")
addWorksheet(wb = wb, sheetName = "Compound metadata")
clean_stans %>%
  select(compound_name, compound_type, formula, stan_rt, stan_mz, 
         kegg_id, polarity, peak_quality) %>%
  writeData(wb = wb, sheet = "Compound metadata")

addWorksheet(wb=wb, sheetName = "README")
data.frame(`README for Ingalls MESOSCOPE data`=c(
  "This file was produced by William Kumler in the Ingalls lab at the University of Washington",
  "This file contains concentrations of compounds in seawater measured via HPLC-MS",
  "in a targeted (selected subset of known compounds) method.",
  "",
  "The first sheet (Sample data) contains the concentrations of various molecules in each sample",
  "taken during the MESOSCOPE cruise in 2017. These values are in nanomoles per",
  "liter of seawater (nM) from the particulate phase (particles collected onto",
  "0.2um Durapore filters).",
  "Missing values are shown as empty cells and infinite values (usually in the standards) are shown as #NUM!",
  "",
  "The second sheet (File metadata) contains information about the samples themselves, including",
  "the depth, time, and station from which they were collected. ",
  "sample_name = Name of the sample as shown in the Sample data Excel sheet",
  "samp_type = Type of sample, whether biological sample (Smp), blank (Blk), pooled (Poo), or standard (Std)",
  "station = Station from which the sample was collected, 1-15 plus L1 and L2",
  "cast = Which cast the sample was taken from",
  "depth = Depth descriptor of where in the water column the sample was collected",
  "abs_depth = Absolute depth. Accounts for differences from DCM and is determined from the CTD itself",
  "rosette = Which Niskin was this sample taken from? Largely incomplete",
  "date_run = The date on which the HPLC-MS run was started",
  "",
  "The third sheet (Compound metadata) contains information about the compounds measured",
  "during this analysis. Not all compounds were detected and passed quality control",
  "so there are more compounds here than found in the Sample data sheet.",
  "compound_name = Name of the compound, as found in the Sample data sheet",
  "compound_type = Broad compound class as determined rather arbitrarily by myself",
  "formula = Molecular formula of the compound. Isotopes are labeled in parentheses",
  "stan_rt = Estimated retention time of the compound, in minutes",
  "stan_mz = Mass-to-charge ratio of the molecule at which it's detected in the mass spec",
  "kegg_id = KEGG (Kyoto Encyclopedia of Genes and Genomes) ID: https://www.genome.jp/kegg/",
  "polarity = The polarity at which the molecule was detected in the mass spec, either 'pos' for positive mode or 'neg' for negative mode",
  "peak_quality = The typical quality of the chromatographic peak. Yes is good, No is bad, and see the Ingalls Standards repo for more details on Problematic compounds",
  "",
  paste("File created on", Sys.Date())
)) %>%
  writeData(wb = wb, sheet = "README")
saveWorkbook(wb, file = "MESOSCOPE_Targeted_nM.xlsx", overwrite = TRUE)
```

```{r write Falkor to Excel}
library(openxlsx)
wb <- createWorkbook(creator = "William Kumler, wkumler@uw.edu")
addWorksheet(wb, sheetName = "Sample data")
falkor_data %>%
  pivot_wider(names_from = filename, values_from = nM) %>%
  writeData(wb = wb, sheet = "Sample data")
freezePane(wb, sheet = "Sample data", firstRow = TRUE, firstCol = TRUE)
addWorksheet(wb, "File metadata")
file_metadata %>%
  filter(cruise=="FK") %>%
  mutate(filename=str_remove(filename, ".mzML")) %>%
  distinct(filename, samp_type, station, cast, depth, abs_depth, rosette, date_run) %>%
  rename(sample_name=filename) %>%
  writeData(wb = wb, sheet = "File metadata")
addWorksheet(wb = wb, sheetName = "Compound metadata")
clean_stans %>%
  select(compound_name, compound_type, formula, stan_rt, stan_mz, 
         kegg_id, polarity, peak_quality) %>%
  writeData(wb = wb, sheet = "Compound metadata")

addWorksheet(wb=wb, sheetName = "README")
data.frame(`README for Ingalls Falkor data`=c(
  "This file was produced by William Kumler in the Ingalls lab at the University of Washington",
  "This file contains concentrations of compounds in seawater measured via HPLC-MS",
  "in a targeted (selected subset of known compounds, manually integrated) method.",
  "",
  "The first sheet (Sample data) contains the concentrations of various molecules in each sample",
  "taken during the Falkor cruise in 2018. These values are in nanomoles per",
  "liter of seawater (nM) from the particulate phase (particles collected onto",
  "0.2um Durapore filters).",
  "Missing values are shown as empty cells and infinite values (usually in the standards) are shown as #NUM!",
  "",
  "The second sheet (File metadata) contains information about the samples themselves, including",
  "the depth, time, and station from which they were collected. ",
  "sample_name = Name of the sample as shown in the Sample data Excel sheet",
  "samp_type = Type of sample, whether biological sample (Smp), blank (Blk), pooled (Poo), or standard (Std)",
  "station = Station from which the sample was collected",
  "cast = Which cast the sample was taken from",
  "depth = Depth descriptor of where in the water column the sample was collected",
  "abs_depth = Absolute depth. Accounts for differences from DCM and is determined from the CTD itself",
  "rosette = Which Niskin was this sample taken from? Largely incomplete",
  "date_run = The date on which the HPLC-MS run was started",
  "",
  "The third sheet (Compound metadata) contains information about the compounds measured",
  "during this analysis. Not all compounds were detected and passed quality control",
  "so there are more compounds here than found in the Sample data sheet.",
  "compound_name = Name of the compound, as found in the Sample data sheet",
  "compound_type = Broad compound class as determined rather arbitrarily by myself",
  "formula = Molecular formula of the compound. Isotopes are labeled in parentheses",
  "stan_rt = Estimated retention time of the compound, in minutes",
  "stan_mz = Mass-to-charge ratio of the molecule at which it's detected in the mass spec",
  "kegg_id = KEGG (Kyoto Encyclopedia of Genes and Genomes) ID: https://www.genome.jp/kegg/",
  "polarity = The polarity at which the molecule was detected in the mass spec, either 'pos' for positive mode or 'neg' for negative mode",
  "peak_quality = The typical quality of the chromatographic peak. Yes is good, No is bad, and see the Ingalls Standards repo for more details on Problematic compounds",
  "",
  paste("File created on", Sys.Date())
)) %>%
  writeData(wb = wb, sheet = "README")
saveWorkbook(wb, file = "Falkor_Targeted_nM.xlsx", overwrite = TRUE)
```
